Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-10 ind_files/02_BD_processes.tex
Minor phrasing changes

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. As I explain in Appendix \ref{App_SDE_FPE}, the Fokker-Planck equation is a `conservation law' for probability, and just mathematically expresses the common-sense observation that the sum (or integral) of probabilities over the entire state space cannot change over time (since it must always equal 1). If the function $G$ is independent of $x$, it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

>>>>

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. As I explain in Appendix \ref{App_SDE_FPE}, the Fokker-Planck equation can be viewed as a `conservation law' for probability that mathematically expresses the common-sense observation that the sum (or integral) of probabilities over the entire state space cannot change over time (since it must always equal 1). If the function $G$ is independent of $x$, it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

### 2023-03-10 ind_files/99_Appendix_A.tex
Minor phrasing changes

\\

@@

where $\div$ is the divergence operator (also denoted $\mathrm{div}$ in some texts). Those familiar with physics should immediately recognize that equation \eqref{app_FPE_ndim_continuity_eqn} is in the form of a so-called `continuity equation' for a conserved quantity. Continuity equations occur in various areas of applied mathematics, most famously in electromagnetism and fluid dynamics. This explains the name `probability current', in analogy with other currents such as electrical current or fluid current. The continuity equation representation also  makes it clear that the Fokker-Planck equation describes the `flow of probability' in the system. In particular, equation \eqref{app_FPE_ndim_continuity_eqn} says that the total probability in the system is `conserved', and is thus simply a mathematical formalization of the common-sense idea that whenever the probability of the system of being in a given state decreases, the probability of it being in some other state must increase (and vice versa).

>>>>

@@

where $\div$ is the divergence operator (also denoted $\mathrm{div}$ in some texts). Those familiar with physics should immediately recognize that equation \eqref{app_FPE_ndim_continuity_eqn} is in the form of a so-called `continuity equation' for a conserved quantity. Continuity equations turn up in various areas of applied mathematics, most famously in electromagnetism (conservation of charge from Maxwell's equations), fluid dynamics (continuity equations for mass of a flowing fluid from the Euler equations), and molecular diffusion (Fick's law). This explains the name `probability current' as an analogy to currents in physics such as electrical current or fluid current. The continuity equation representation also  makes it clear that the Fokker-Planck equation describes the `flow of probability' in the system. In particular, equation \eqref{app_FPE_ndim_continuity_eqn} says that the total probability in the system is `conserved', and is thus simply a mathematical formalization of the common-sense idea that whenever the probability of the system of being in a given state decreases, the probability of it being in some other state must increase (and vice versa).
